---
title: "Visualizing crime in New Haven"
author: "Dan Weinberger"
date: "April 27, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgdal)
library(raster)
library(shiny)
library(lubridate)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(shinydashboard)

source('./R/crimeplot.fun.R')
source('./R/app.R')

```

## New Haven Crime Data
These data were provided by Mark Abraham at DataHaven (info@ctdatahaven.org ). *These data are not for public dissemination. If you would like to use the data outside of class, please contact Mark directly*. These data on crime in New Haven cover the period 2013-2015 are collected by the New Haven police department Each crime receives a Uniform Crime Reporting (UCR) code--basically the ICD of the police world. The full dataset is extremely detailed, with every crime reported in the city, with information on address, latitude and longitude, census tract, time and date of the event, and the crime committed. I have gone through and made some modifications to the data: I dropped variables we don't need, I am only including a subset of the crimes, and I am 'jittering' the coordinates of each event by a random amount in each direction for privacy purposes.  

## Mapping
Today we will map the data in 2 ways: by point and by neighborhood
First we need to import the shape file into R using the shapefile() function. This has neighborhood boundaries for all cities in Connecticut and was donwloaded from the Zillow website.

```{r}
shp1 <- readRDS('./Data/nh.shp.save.rds')
shp.nhv<-shp1[shp1$City=="New Haven",] #subset to only include New Haven

shp.nhv2 <- fortify(shp.nhv)

```

## read in the data
```{r}
n2<-readRDS('./Data/nh.crime.cleaned.rds')
```


## Explore the data
```{r}
str(n2)
View(n2)
```


Just like we did earlier, we are going to map the points This is all of the crimes in our dataset. Play around with alpha to get a good level of transparency

```{r}
par(mfrow=c(1,1))
red.trans<-rgb(1, #Red (between 0 and 1)
               0, #Green (between 0 and 1)
               0, #Blue (between  and 1)
               alpha=1  #between 0 and 1--controls transparency
               )


p1a <- ggplot(n2, aes(x=x.jit, y=y.jit)) +
  geom_point( col='red',
              alpha=0.1,
              cex=0.5, #Size of the markers
              pch=16) +
    theme_void()+ #turns off axes
  ggtitle('Location of Crimes')
  
p1 <-  p1a+ 
      geom_polygon(data=shp.nhv2, aes(x=long, y=lat, group=group), fill=NA, colour='black', alpha=0) 

p1
```

## Just look at a subset of the crimes. 
View the vector of crimes in our database, and select which we want to view
```{r}
unique(n2$NIBRS_Offe)
sub1<-n2[n2$NIBRS_Offe=='DISORDERLY CONDUCT',]
```

```{r}
p1 <- crimeplot.fun(select.crimes="DISORDERLY CONDUCT",
                    opacity1=0.1)
```

## Let's also look at the time of the crime
```{r}
sub1 <- n2[n2$NIBRS_Offe=='DISORDERLY CONDUCT',]
hist(sub1$Hr)

p2 <- ggplot(sub1, aes(x=Hr)) + 
  geom_histogram(binwidth=1)+
  theme_classic()+
  ggtitle('Time of Crime') +
  xlab('Time of Day')
```

#And finally the date--can aggregate by day or week or month
Play around with aggregation level
```{r}
sub1<-n2[n2$NIBRS_Offe=='DISORDERLY CONDUCT',]
```

Tell R the date is a date variable
```{r}
  sub1$date1<-as.Date(sub1$Date_Occur, '%m/%d/%Y')
```

Round date to 'day' or 'week'
```{r}
    sub1$date.week<-floor_date( sub1$date1, unit='day')
```

Use aggregate to count number of incidents per unit of time
```{r}
sub2 <- sub1 %>%
    group_by(date.week) %>%
    summarize(N_crimes=n())
sub2 <- sub2[!is.na(sub2$date.week),]
```

```{r}
p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
  geom_line() +
  theme_classic() +
  ylab('N crimes')+
  xlab('Date')
   ##hmm..what could have caused the spike in disorderly conduct mid-March 2014?
    p3
```

#lets combine them
Patchwork package makes this easy
```{r, fig.width=5, fig.height=4}

(p1+ plot_spacer())/( p2+p3)
```

## Making an interactive plot
RShiny (shiny package) provides a way to transform your R code into interactive visualizations with relatively little effort. Wow your employers with your wizardry... 
There is a great gallery of examples with code you can modify here: https://shiny.rstudio.com/gallery/

There are 2 components to a Shiny app: a user interface (UI) and a server. The UI controls how everything is laid out, how it looks, and what options are available to the user. The server is the business end and runs analyses (if any) generates the plots, etc.

#Let's figure out the layout.
first the UI:
```{r}
ui <- pageWithSidebar(

  # App title ----
  headerPanel("YOUR TITLE HERE"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                c('Crime 1', 'Crime 2'))
  ),

  # Main panel for displaying outputs ----
  mainPanel(

      )
)
```

Then the 'server'...doesn't do anything yet
```{r}
server <- function(input, output) {
  
}
```

Then Call the app
```{r}
shinyApp(ui, server)
```

## Add a slider for opacity of the points

```{r}
ui <- pageWithSidebar(

  # App title ----
  headerPanel("YOUR TITLE HERE"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                c('Crime 1', 'Crime 2')),
                
      sliderInput("opacity", "Opacity:", 
                min=0, max=1, value=0.02)
  ),

  # Main panel for displaying outputs ----
  mainPanel(

      )
)
```

Now the server
```{r}
server <- function(input, output) {
  
}
```

Call the app
```{r}
shinyApp(ui, server)
```

## Have the dropdown menu fill with the actual crimes
And change the title
```{r}
crime.vector<- unique(n2$NIBRS_Offe) #generate a vector with all of the crimes listed
ui <- pageWithSidebar(

  # App title ----
  headerPanel("New Haven Crime Statistics"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                crime.vector),
                
      sliderInput("opacity", "Opacity:", 
                min=0, max=1, value=0.02)
  ),

  # Main panel for displaying outputs ----
  mainPanel(

      )
)

server <- function(input, output) {
  
}
shinyApp(ui, server)
```

## Ok it looks good, now let's make it do something
To have it do anything, we need to set up the server code. We are basically just going to cut and paste the map code from above and make some of the variables dynamic. In the server, you access the input values that you selected on the UI using input$VARNAME, where varname is the name of the slider/dropdown/etc. 

It is saving an output called "crimePlot" that can be accessed by the UI by saving the plot to output$crimePlot. 
To make the output appear, add the the plotOutput("crimePlot") statement in the mainPanel

```{r}
crime.vector<-unique(n2$NIBRS_Offe) 
ui <- pageWithSidebar(

  # App title ----
  headerPanel("New Haven Crime Statistics"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                crime.vector),
                
      sliderInput("opacity", "Opacity:", 
                min=0, max=1, value=0.02)
  ),

  # Main panel for displaying outputs ----
  mainPanel(
    plotOutput("crimePlot")
      )
)

server <- function(input, output) {
  
  #The inputs from the dropdown and slider get saved in an object called 'input'. We can refer to these objects anywhere between the curly brackets  below to dynamically change the plots
  output$crimePlot <- renderPlot({ #Saves an object called 'crimePlot' that we can plot in the UI
    
    #Subset the full dataset based on the dropdown menu selection
    sub1<-n2[n2$NIBRS_Offe==input$crime,]
   
    #Make a histogram to be plotted
    p2 <- ggplot(sub1, aes(x=Hr)) + 
      geom_histogram(binwidth=1)+
      theme_classic()+
      ggtitle('Time of Crime') +
      xlab('Time of Day')
    p2
    })
}
shinyApp(ui, server)
```

## Same thing with a map
```{r}
crime.vector<-unique(n2$NIBRS_Offe) 
ui <- pageWithSidebar(

  # App title ----
  headerPanel("New Haven Crime Statistics"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                crime.vector),
                
      sliderInput("opacity", "Opacity:", 
                min=0, max=1, value=0.02)
  ),

  # Main panel for displaying outputs ----
  mainPanel(
    plotOutput("crimePlot")

      )
)

server <- function(input, output) {
  
  #The inputs from the dropdown and slider get saved in an object called 'input'. We can refer to these objects anywhere between the curly brackets  below to dynamically change the plots
  output$crimePlot <- renderPlot({ #Saves an object called 'crimePlot' that we can plot in the UI
    
    #Subset the full dataset based on the dropdown menu selection
    #Set the opacity level based on the slider value

    #All of this is the same as what we had before
    p1 <- crimeplot.fun(select.crimes=input$crime,
                    opacity1=input$opacity)
    p1
    })
  
}
shinyApp(ui, server)
```


#And let's just get fancy and combine together histograms for time and date and map...
```{r}
crime.vector<-unique(n2$NIBRS_Offe) 
ui <- pageWithSidebar(

  # App title ----
  headerPanel("New Haven Crime Statistics"),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
  sidebarPanel(
      # Input: Selector for variable to plot against mpg ----
      selectInput("crime", "Select Crime:", 
                crime.vector),
      selectInput("time.agg", "Time aggregation:", 
                c('day','week','month')),
      sliderInput("opacity", "Opacity:", 
                min=0, max=1, value=0.02)
  ),

  # Main panel for displaying outputs ----
  mainPanel(
    plotOutput("crimePlot")
      )
)

server <- function(input, output) {
  
  #The inputs from the dropdown and slider get saved in an object called 'input'. We can refer to these objects anywhere between the curly brackets  below to dynamically change the plots
  output$crimePlot <- renderPlot({ #Saves an object called 'crimePlot' that we can plot in the UI
    
    #Subset the full dataset based on the dropdown menu selection
    sub1<-n2[n2$NIBRS_Offe==input$crime,]
    
    #Set the opacity level based on the slider value
  p1 <- crimeplot.fun(select.crimes=input$crime,
                    opacity1=input$opacity)
  
  p2 <- ggplot(sub1, aes(x=Hr)) + 
      geom_histogram(binwidth=1)+
      theme_classic()+
      ggtitle('Time of Crime') +
      xlab('Time of Day')
  
    sub1$Date_Occur <- as.Date(sub1$Date_Occur, '%m/%d/%Y')
  sub1$date.week <- floor_date( sub1$Date_Occur, unit=input$time.agg)

  sub2 <- sub1 %>%
    group_by(date.week) %>%
    summarize(N_crimes=n())
sub2 <- sub2[!is.na(sub2$date.week),]

   p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
  geom_line() +
  theme_classic() +
  ylab('N crimes')+
  xlab('Date')
   
(p1+ plot_spacer())/( p2+p3)

    
      })
}
shinyApp(ui, server)
```

To get a nicer layout, you can use the ShinyDashboard app. This has all sorts of advanced layouts, like tabs. You can open app.R and launch from there
```{r}
runApp(app2)

```

