---
title: "Visualizing crime in New Haven"
author: "Dan Weinberger"
date: "Feb 15, 2023"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgdal)
library(raster)
library(shiny)
library(lubridate)
library(ggplot2)
library(plotly)
library(patchwork)
library(RColorBrewer)
library(shinydashboard)
library(dplyr)

source('./R/crimeplot.fun.R')
source('./app.R')

```

## New Haven Crime Data

These data were provided by Mark Abraham at DataHaven
([info\@ctdatahaven.org](mailto:info@ctdatahaven.org){.email} ).
***These data are not for public dissemination. If you would like to use
the data outside of class, please contact Mark directly***. These data
on crime in New Haven cover the period 2013-2015 are collected by the
New Haven police department Each crime receives a Uniform Crime
Reporting (UCR) code--basically the ICD of the police world. The full
dataset is extremely detailed, with every crime reported in the city,
with information on address, latitude and longitude, census tract, time
and date of the event, and the crime committed. I have gone through and
made some modifications to the data: I dropped variables we don't need,
I am only including a subset of the crimes, and I am 'jittering' the
coordinates of each event by a random amount in each direction for
privacy purposes.

## Mapping

Today we will map the data in 2 ways: by point and by neighborhood First
we need to import the shape file into R using the shapefile() function.
This has neighborhood boundaries for all cities in Connecticut and was
donwloaded from the Zillow website.

```{r}
shp1 <- readRDS('./Data/nh.shp.save.rds')
shp.nhv<-shp1[shp1$City=="New Haven",] #subset to only include New Haven

shp.nhv2 <- fortify(shp.nhv)

```

## read in the data

```{r}
n2<-readRDS('./Data/nh.crime.cleaned.rds')
```

## Explore the data

```{r}
str(n2)
View(n2)
```

Just like we did earlier, we are going to map the points This is all of
the crimes in our dataset. Play around with alpha to get a good level of
transparency

```{r}
p1a <- ggplot(n2, aes(x=x.jit, y=y.jit)) +
  geom_point( col='red', #color of the points
              alpha=0.1, #Conrols transparency of the points
              cex=0.5, #Size of the markers
              pch=16) + #shape of the points
    theme_void()+ #turns off axes
  ggtitle('Location of Crimes')
  
p1 <-  p1a+ 
      geom_polygon(data=shp.nhv2, aes(x=long, y=lat, group=group), fill=NA, colour='black', alpha=0) 

p1
```

## Just look at a subset of the crimes.

View the vector of crimes in our database, and select which we want to
view

```{r}
unique(n2$NIBRS_Offe)
sub1<-n2 %>%
  filter(NIBRS_Offe=='DISORDERLY CONDUCT') 
```

```{r}
sub1 <- n2 %>%
    filter(NIBRS_Offe=='DISORDERLY CONDUCT') 

p1 <- crimeplot.fun(ds=sub1,
                    opacity1=0.1)

p1
```

## Let's also look at the time of the crime

```{r}
sub1 <- n2[n2$NIBRS_Offe=='DISORDERLY CONDUCT',]

p2 <- ggplot(sub1, aes(x=Hr)) + 
  geom_histogram(binwidth=1)+
  theme_classic()+
  ggtitle('Time of Crime') +
  xlab('Time of Day')

p2
```

#And finally the date--can aggregate by day or week or month Play around
with aggregation level

```{r}
sub1<-n2 %>%
  filter(NIBRS_Offe=='DISORDERLY CONDUCT')

```

Tell R the date is a date variable

```{r}
sub1 <- sub1 %>%
  mutate(date1=as.Date(Date_Occur, '%m/%d/%Y'))

```

Round date to 'day' or 'week'

```{r}
sub1 <- sub1 %>%
  mutate(date.week=floor_date( date1, unit='day'))
```

Use aggregate to count number of incidents per unit of time

```{r}
sub2 <- sub1 %>%
    group_by(date.week) %>%
    summarize(N_crimes=n()) %>%
  filter(!is.na(date.week))
```

```{r}
p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
  geom_line() +
  theme_classic() +
  ylab('N crimes')+
  xlab('Date')
   ##hmm..what could have caused the spike in disorderly conduct mid-March 2014?
    p3
```

Make it interactive using plotly

```{r}
ggplotly(p3)
```

#lets combine them Patchwork package makes this easy

```{r, fig.width=5, fig.height=4}

(p1+ plot_spacer())/( p2+p3)
```

```{r}
crime.vector<-unique(n2$NIBRS_Offe) 
crime.vector
```

## Making an interactive plot

RShiny (shiny package) provides a way to transform your R code into
interactive visualizations with relatively little effort. Wow your
employers with your wizardry... There is a great gallery of examples
with code you can modify here: <https://shiny.rstudio.com/gallery/> We
are going to use an add on to Shiny called ShinyDashboard, which makes
things look even nicer.

There are 2 components to a Shiny app: a user interface (UI) and a
server. The UI controls how everything is laid out, how it looks, and
what options are available to the user. The server is the business end
and runs analyses (if any) generates the plots, etc.

#Let's figure out the layout. first the UI:

```{r}
ui <- dashboardPage(
    
    # App title ----

    dashboardHeader(title="YOUR TITLE HERE",titleWidth=500),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    c('Crime 1', 'Crime 2'))
        ),
        
  # Main panel for displaying outputs ----
 dashboardBody(
     fluidRow(
     )
      )
  )

```

Then the 'server'...doesn't do anything yet

```{r}
server <- function(input, output) {
  
}
```

Then Call the app

```{r}
shinyApp(ui, server)
```

## Add a slider for opacity of the points

```{r}
ui <- dashboardPage(
    
    # App title ----

    dashboardHeader(title="YOUR TITLE HERE",titleWidth=500),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    c('Crime 1', 'Crime 2')),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
        ),

        
  # Main panel for displaying outputs ----
 dashboardBody(
     fluidRow(
     )
      )
  )
```

Now the server

```{r}
server <- function(input, output) {
  
}
```

Call the app

```{r}
shinyApp(ui, server)
```

## Have the dropdown menu fill with the actual crimes

And change the title

```{r}
ui <- dashboardPage(
    
    # App title ----

    dashboardHeader(title="New Haven Crime Statistics",titleWidth=500),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
        ),

        
  # Main panel for displaying outputs ----
 dashboardBody(
     fluidRow(
     )
      )
  )

server <- function(input, output) {
  
}
shinyApp(ui, server)
```

## Ok it looks good, now let's make it do something

To have it do anything, we need to set up the server code. We are
basically just going to cut and paste the map code from above and make
some of the variables dynamic. In the server, you access the input
values that you selected on the UI using input\$VARNAME, where varname
is the name of the slider/dropdown/etc.

It is saving an output called "plot_time" that can be accessed by the UI
by saving the plot to output\$plot_time To make the output appear, add
the the plotOutput("plot_time") statement in the mainPanel

```{r}
ui <- dashboardPage(
    
    # App title ----

    dashboardHeader(title="New Haven Crime Statistics",titleWidth=500),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
        ),

        
  # Main panel for displaying outputs ----
 dashboardBody(
     fluidRow(
      box(tabPanel("Time", plotOutput("plot_time")), width=12)
        )
      )
  )

server <- function(input, output) {
    output$plot_time = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      p2 <- ggplot(sub1, aes(x=Hr)) + 
        geom_histogram(binwidth=1)+
        theme_classic()+
       # ggtitle(paste('Time of Crime for ', input$crime)) +
        xlab('Time of Day')
      p2
    })
}
shinyApp(ui, server)
```

## Same thing with a map

see the slider now works...

```{r}
ui <- dashboardPage(
    
    # App title ----

    dashboardHeader(title="New Haven Crime Statistics",titleWidth=500),

  # Sidebar panel for inputs--these can be sliders, dropdown menus, etc
dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
        ),

        
  # Main panel for displaying outputs ----
 dashboardBody(
     fluidRow(
          box(tabPanel("Map", plotOutput("map1")), width=12),
        )
      )
  )

server <- function(input, output) {
   output$map1 = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      
      p1 <- crimeplot.fun(ds=sub1,
                          opacity1=input$opacity)
      p1
    })
}
shinyApp(ui, server)
```

#And let's just get fancy and combine together histograms for time and
date and map...

```{r}
 ui = dashboardPage(
    
    dashboardHeader(title = "New Haven Crime Statistics",titleWidth=500),
    
    dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        selectInput("time.agg", "Time aggregation:", 
                    c('day','week','month')),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
      ),
    dashboardBody(
      fluidRow(
                    box(tabPanel("Map", plotOutput("map1")), width=4),
                    box(tabPanel("Time", plotOutput("plot_time")), width=4),
                    box(tabPanel("Date", plotOutput("plot_date")), width=4)
      )
    )
  )
  
#SERVER
server = function(input, output) {
    output$map1 = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      
      p1 <- crimeplot.fun(ds=sub1,
                          opacity1=input$opacity)
      p1
    })
    
    output$plot_time = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      p2 <- ggplot(sub1, aes(x=Hr)) + 
        geom_histogram(binwidth=1)+
        theme_classic()+
        ggtitle('Time of Crime') +
        xlab('Time of Day')
      p2
    })
    
    output$plot_date = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      sub1$Date_Occur <- as.Date(sub1$Date_Occur, '%m/%d/%Y')
      sub1$date.week <- floor_date( sub1$Date_Occur, unit=input$time.agg)
      
      sub2 <- sub1 %>%
        group_by(date.week) %>%
        summarize(N_crimes=n())
      sub2 <- sub2[!is.na(sub2$date.week),]
      
      p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
        geom_line() +
        theme_classic() +
        ylab('N crimes')+
        xlab('Date')
      p3
    })
      
  }
shinyApp(ui, server)
```

## OR we could try a different layout

```{r}
 ui = dashboardPage(
    
    dashboardHeader(title = "New Haven Crime Statistics",titleWidth=500),
    
    dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        selectInput("time.agg", "Time aggregation:", 
                    c('day','week','month')),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
      ),
    dashboardBody(
                       box(tabPanel("Map", plotOutput("map1"))),
                    box(tabPanel("Time", plotOutput("plot_time"))),
                    box(tabPanel("Date", plotOutput("plot_date")))
      )
  )
  
#SERVER
server = function(input, output) {
    output$map1 = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      
      p1 <- crimeplot.fun(ds=sub1,
                          opacity1=input$opacity)
      p1
    })
    
    output$plot_time = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      p2 <- ggplot(sub1, aes(x=Hr)) + 
        geom_histogram(binwidth=1)+
        theme_classic()+
        ggtitle('Time of Crime') +
        xlab('Time of Day')
      p2
    })
    
    output$plot_date = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      sub1$Date_Occur <- as.Date(sub1$Date_Occur, '%m/%d/%Y')
      sub1$date.week <- floor_date( sub1$Date_Occur, unit=input$time.agg)
      
      sub2 <- sub1 %>%
        group_by(date.week) %>%
        summarize(N_crimes=n())
      sub2 <- sub2[!is.na(sub2$date.week),]
      
      p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
        geom_line() +
        theme_classic() +
        ylab('N crimes')+
        xlab('Date')
      p3
    })
      
  }
shinyApp(ui, server)
```

## Add a value box

```{r}
 ui = dashboardPage(
    
    dashboardHeader(title = "New Haven Crime Statistics",titleWidth=500),
    
    dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        selectInput("time.agg", "Time aggregation:", 
                    c('day','week','month')),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
      ),
    dashboardBody(
                       box(tabPanel("Map", plotOutput("map1"))),
                    box(tabPanel("Time", plotOutput("plot_time"))),
                    box(tabPanel("Date", plotOutput("plot_date"))),
      # Dynamic valueBoxes
      valueBoxOutput("TallyBox"),
      )
  )
  
#SERVER
server = function(input, output) {
    output$map1 = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      
      p1 <- crimeplot.fun(ds=sub1,
                          opacity1=input$opacity)
      p1
    })
    
    output$plot_time = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      p2 <- ggplot(sub1, aes(x=Hr)) + 
        geom_histogram(binwidth=1)+
        theme_classic()+
        ggtitle('Time of Crime') +
        xlab('Time of Day')
      p2
    })
    
    output$plot_date = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      sub1$Date_Occur <- as.Date(sub1$Date_Occur, '%m/%d/%Y')
      sub1$date.week <- floor_date( sub1$Date_Occur, unit=input$time.agg)
      
      sub2 <- sub1 %>%
        group_by(date.week) %>%
        summarize(N_crimes=n())
      sub2 <- sub2[!is.na(sub2$date.week),]
      
      p3 <- ggplot(sub2, aes(x=date.week, y=N_crimes))+
        geom_line() +
        theme_classic() +
        ylab('N crimes')+
        xlab('Date')
      p3
    })
    
    #Tally box
     output$TallyBox <- renderValueBox({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
    valueBox(
      paste0(nrow(sub1)), input$crime, icon = icon("list"),
      color = "purple"
    )
  })
      
  }
shinyApp(ui, server)
```

## Make plots interactive with plotly

```{r}
 ui = dashboardPage(
    
    dashboardHeader(title = "New Haven Crime Statistics",titleWidth=500),
    
    dashboardSidebar(
        selectInput("crime", "Select Crime:", 
                    crime.vector),
        selectInput("time.agg", "Time aggregation:", 
                    c('day','week','month')),
        sliderInput("opacity", "Opacity:", 
                    min=0, max=1, value=0.02)
      ),
    dashboardBody(
      fluidRow(
      box(tabPanel("Map", plotOutput("map1"))),
      box(tabPanel("Time", plotOutput("plot_time"))),
      box(tabPanel("Date", plotlyOutput("plot_date"))),
      # Dynamic valueBoxes
      valueBoxOutput("TallyBox"),
      ))
  )
  
#SERVER
server = function(input, output) {
    output$map1 = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      
      p1 <- crimeplot.fun(ds=sub1,
                          opacity1=input$opacity)
      p1
    })
    
    output$plot_time = renderPlot({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      p2 <- ggplot(sub1, aes(x=Hr)) + 
        geom_histogram(binwidth=1)+
        theme_classic()+
        ggtitle('Time of Crime') +
        xlab('Time of Day')
      p2
    })
    
    output$plot_date = renderPlotly({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
      sub1$Date_Occur <- as.Date(sub1$Date_Occur, '%m/%d/%Y')
      sub1$date.week <- floor_date( sub1$Date_Occur, unit=input$time.agg)
      
      sub2 <- sub1 %>%
        group_by(date.week) %>%
        summarize(N_crimes=n())
      sub2 <- sub2[!is.na(sub2$date.week),]
      
      p3 <- ggplotly(ggplot(sub2, aes(x=date.week, y=N_crimes))+
        geom_line() +
        theme_classic() +
        ylab('N crimes')+
        xlab('Date')
      )
      p3
    })
    
    #Tally box
     output$TallyBox <- renderValueBox({
      sub1<-n2[n2$NIBRS_Offe==input$crime,]
    valueBox(
      paste0(nrow(sub1)), input$crime, icon = icon("list"),
      color = "purple"
    )
  })
      
  }
shinyApp(ui, server)
```
